Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-12-02 Manuscript/manuscriptPGPE.tex
Update manuscript

Decrease in headgroup order parameters of PC lipids upon addition of charges is
usually captured qualitatitely despite inaccurate lipid conformational ensembles in the models \cite{catte16},
but situation with PS lipids is more complicated \cite{antila19,meclr20}.
Here we observe in figure \ref{changesWITHCaClPG} that
Lipid17 and Slipids force fields correctly capture the PG $\beta$-carbon order parameter response to CaCl$_2$
even thought the calcium binding affinity was overestimated.
%based on the comparison of PC headgroup order parameter changes with experiments.
%While applying ECC to Lipid17 improved the PC headgroup order parameter response and binding affinity,
%The response of PG $\beta$-carbon order parameter to calcium is too weak in CHARMM36 and ECC-lipid17 force fields,
%but decrease of the order parameter is associated with decreased P-N vector angle in all force fields.
Analysis of the changes in PG conformational ensemble from these models suggest that
calcium is affecting also to the headgroup glycerol conformations
(Figs \ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}),
in contrast to PC lipids where only conformations near phosphate were affected.
%dihedral angle distribution around O$_\alpha$-C$_\alpha$ bond is affected by calcium binding (Fig. ??).
On the other hand, the upward tilting of the headgroup dipole is weaker
%Even thought the change in headgroup orientation upon addition of calcium is smaller
in PG than in PC headgroup, possible due to the compenating effects from the changes in
phosphate and glycerol regions. 
However, none of the simulations captures and calcium binding affinity and conformational ensemble of PG lipids
simulationeously, and differences between force fields are observed for the response of PG $\alpha$ carbon to
the added calcium for which experimental data is not available.
Therefore, more accurate force fields are required for the solid analysis of PG conformations enembles in different ionic conditions.

@@

\subsection{Protein bound lipid conformations }
\begin{figure*}[!h]
  \centering
  \includegraphics[width=8.0cm]{../Figs/DIHEDRALSPCfromPDB.eps}
  \includegraphics[width=8.0cm]{../Figs/DIHEDRALSPEfromPDB.eps}
  \includegraphics[width=8.0cm]{../Figs/DIHEDRALSPGfromPDB.eps}
  \includegraphics[width=8.0cm]{../Figs/DIHEDRALSPSfromPDB.eps}
  \caption{\label{structures}
    Dihedral distributions from simulations and lipid structures in PDB.
  }
\end{figure*}

\clearpage
\section{Conclusions}

>>>>

Decrease in headgroup order parameters of PC lipids upon addition of charges is
usually qualitatitely captured in simulations despite the inaccuries in lipid conformational ensembles \cite{catte16},
but situation with PS lipids is more complicated \cite{antila19,meclr20}.
Here, Lipid17 and Slipids force fields correctly capture the PG $\beta$-carbon order parameter response to CaCl$_2$
in figure \ref{changesWITHCaClPG} even thought the calcium binding affinity was overestimated.
%based on the comparison of PC headgroup order parameter changes with experiments.
%While applying ECC to Lipid17 improved the PC headgroup order parameter response and binding affinity,
%The response of PG $\beta$-carbon order parameter to calcium is too weak in CHARMM36 and ECC-lipid17 force fields,
%but decrease of the order parameter is associated with decreased P-N vector angle in all force fields.
Heavy atom dihedral angle distributions calculated from these simulations
%the changes in PG conformational ensemble from these models
suggest that also the headgroup glycerol conformations in PG are affected by calcium
(Figs \ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}),
in contrast to PC lipids where only conformations near phosphate were affected.
%dihedral angle distribution around O$_\alpha$-C$_\alpha$ bond is affected by calcium binding (Fig. ??).
On the other hand, the upward tilting of the headgroup dipole is weaker
%Even thought the change in headgroup orientation upon addition of calcium is smaller
in PG than in PC headgroup, possible due to the compensating effects from the changes in
phosphate and glycerol regions. 
However, none of the simulations captures and calcium binding affinity and conformational ensemble of PG lipids
simultaneously, and experimental data to evaluate 
%differences between force fields are observed for
the response of $\alpha$-carbon order parameters to
the added calcium in PG is not available.
Therefore, more accurate force fields are required for the solid analysis of PG conformations enembles in different ionic conditions.

@@

Despite the limited cabability of simulations to interpret some experimental data due to
the inaccuries in force fields, we can conclude that the experimentally observed changes
in headgroup order parameters upon addition of charges into bilayer arise from relatively
small changes in conformational ensembles. These changes can be characterised by
mild changes in dihedral angle distributions, rather than restriction of lipids into
fixed conformations. Therefore, lipid headgroups remain in disordered state sampling large
space of different conformations, thereby being able to interact with different molecules
in various ways, also when charged molecules are bound to membranes.

\subsection{Protein bound lipid conformations}

While our results in previous sections suggest that lipid headgroups
sample large conformational space in liquid lamellar phase, lipids
can bound to proteins in fixed conformations which can be then detected
using crystallography or cryo-EM. Such protein bound lipid conformations are
available within protein structures deposited in The Protein Data Bank (PDB)
available at \url{http://www.rcsb.org/} \cite{berman00}. With our criterias described
in the methods, we found
?? conformations of PC lipids,
?? conformations of PE lipids,
?? conformations of PG lipids, and
?? conformations of PS lipids from PDB.

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/DIHEDRALSALLfromPDB.eps}
%  \includegraphics[width=8.0cm]{../Figs/DIHEDRALSPEfromPDB.eps}
%  \includegraphics[width=8.0cm]{../Figs/DIHEDRALSPGfromPDB.eps}
%  \includegraphics[width=8.0cm]{../Figs/DIHEDRALSPSfromPDB.eps}
  \caption{\label{dihedralsFROMpdb}
    Dihedral distributions from simulations and lipid structures in PDB.
  }
\end{figure}

To analyze the relation of lipid headgroup conformational ensembles between
liquid lamellar and protein bound states, we calculated the heavy atom dihedral angle
distributions also from the protein conformations found from the PDB.
The results in figure \ref{dihedralsFROMpdb} do not reveal major differences
between different lipid headgroups when bound to proteins.
Some preference for trans state in C$_\alpha$-C$_\beta$ bond of PC lipids, and
for positive angles in O$_\alpha$-C$_\alpha$ and P-O$_\alpha$ of PS lipids with respect
to other lipids may be present in the data, but more statistics is required for solid
conclusions. Therefore, the differences in conformational ensembles between different
lipids observed in liquid lamellar state are not clearly visible in the protein bound states.

Almost all dihedral angles are found from the protein bounds states for all dihedrals
except for P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P, which seem to avoid
the cis conformations. The wide range of available dihedrals is similar to our results
in liquid lamellar state, except for O$_\alpha$-C$_\alpha$-C$_\beta$-/C$_\gamma$ and
g$_1$-g$_2$-g$_3$-O$_{g_3}$ dihedrals which seem to be more restricted in the liquid
lamellar phase (Fig. \ref{structures}). This suggests that lipids compromize their
preferred conformations when bound to proteins to minimize the intermolecular interactions
with proteins.

\section{Conclusions}

We have measured the C-H bond order parameters with the signs for headgroup and glycerol
backbone regions of the most most abundant biological phopholipids, PC, PE, PG and PS.
Combining this experimental data with the large amount of simulation data collected 
using the NMRlipids open collaboration, we were able to resolve the differences between
conformational ensembles of lipid headgroup leading to the differences observed in NMR experiments.

Our results indicate that lipid headgroups are flexible and sample large conformational space
in biologically relevant liquid lamellar state, also in membranes containing charged molecules,
which have the largest affect on lipid headgroups in NMR experiments.
Therefore, lipids can bind to proteins and other biomolecules in multiple different conformations,
as indeed observed in the analysis of protein bound lipid conformations available from PDB.
The weak correlation between lipid structures in liquid lamellar phase and protein bound state
suggestst that the selective lipid binding to proteins is dominated by the intermolecular lipid-protein
interactions, and the differences in conformational ensembles between different lipid types
play only a minor role.

Our results pave the way to understand how lipids regulate membrane protein function.
For example, the resolved lipid conformational ensembles in liquid lamellar phase and
realistic MD simulations enables accurate analysis of lipid-protein interactions energetics.
Furthermore, our results demonstrate how NMR data can be combined with the large amount of
indexed MD simulation data to find the most realistic conformational ensembles of biomolecules
in membrane environment. This paves the way toward standardized methods to resolve the quality
evaluated conformational ensembles of disordered biomolecules in membrane environment.
In the NMRlipids project we aim to build a general databank of quality evaluated
conformational ensembles of lipids and other disordered biomolecules based primarily on
MD simulations and NMR data.

### 2020-12-02 Manuscript/manuscriptPGPEsuppl.tex
Update manuscript

\clearpage
\section{Comparison of headgroup order parameters from different force fields to experiments}
The poor performance of headgroup order parameters in Berger model can be probably explained by
ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04}, which is a typical feature
for Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
The poor performance of glycerol backbone of Slipids simulations is systemically observed also
for other lipids in previous studies \cite{botan15,antila19}.
\todo{Should we comment more the relative quality of different force fields and/or make the subjective force field ranking figures?
https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/8}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPE.pdf}
  \caption{\label{HGorderParametersPE}
    The headgroup and glycerol backbone order parameters of PE lipids
    from experiments (POPE and signs this work, DPPE from Ref.~\citenum{seelig76}
    and E.coliPE from Ref.~\citenum{gally81}) and simulations with different force fields.
  }
  \todo{This should be clarified as in NMRlipidsI and error bars should be added.
    Probably larger error bars for united atom models based on the report by Fuchs et al.
  } \\
\end{figure}

@@

\pagebreak

\subsection{PC headgroup interactions with PE and PG}
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of POPC headgroup order parameters with increasing amount of POPE (left) and POPG (right) in bilayer
    from experiments \cite{scherer87,macdonald87} and simulations with different force fields.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{P-N vector angles from Berger-OPLS and GROMOS-CKP simulations are yet to be analyzed.}
\end{figure*}

In experiments, the PC headgroup order parameters increase
with the addition of negatively charged PG or PS lipids, but
are not affected by the addition of zwitterionic PE and SM lipids
or cholesterol (Fig.~\ref{HGorderparametersPCvsPEPG}).
This can be explained by the electrometer concept, which suggests that the 
headgroup dipole tilts more parallel to the membrane plane upon addition of
negative charge to the membrane~\cite{seelig87, scherer87,antila18}.
The response of PC headgroup order parameters to PE by the tested CHARMM36 and
Berger-OPLS force fields, although CHARMM36 slightly overestimates the changes (Fig.~\ref{HGorderparametersPCvsPEPG}).
The good performance of Berger-OPLS simulations is notable because the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/Höltje force field in our previous work \cite{botan15}.
\todo{This is text by P. Fuchs, copied from the blog. \\
Area results in nm$^2$, the error is $<=$ 0.003 nm$^2$ \\
- pure POPC \\
CHARMM36: 0.624 \\
Berger : 0.649 \\
- POPC/POPE 50:50 \\
CHARMM36 : POPC 0.609, POPE 0.557 \\
Berger-hacked: POPC 0.637, POPE 0.632 \\
----- \\
One can see that CHARMM 36 predicts a drop in the area on going from pure POPC to POPC/POPE 50:50. This means that POPC pack tightly to POPE. 
In contrast, the values for Berger are not that changed. 
The POPE value predicted by CHARMM 36 (in the mixture POPC/POPE 50:50) is much smaller than that predicted by Berger.\\
-------\\
The experimental acyl chain order parameters for POPE \cite{pare98} seem larger than reported for POPC \cite{ferreira13},
which supports the more condensed PE bilayer.
This is interesting, but not exactly the core message of the manuscript.
Maybe we should mention this very briefly? For example, we could just report the areas per lipid (without distinguishing PC and PE)
and mention the difference between CHARMM36 and Berger. I have opened an issue for this: https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/7
}

>>>>

\clearpage
\section{Comparison of headgroup order parameters from different force fields to experiments}

The quality of PE and PG headgroup conformational ensembles in different simulations against NMR experiments
is evaluated in figures \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}
using C-H bond order parameters as in our previous studies for PC and PS lipids \cite{botan15,antila19}.
Conclusions are the same for all lipids: None of the force fields correctly captures the lipid headgroup
conformational ensembles, but CHARMM36 gives results closest to experiments.

%The poor performance of headgroup order parameters in Berger model can be probably explained by
%ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04}, which is a typical feature
%for Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
%The poor performance of glycerol backbone of Slipids simulations is systemically observed also
%for other lipids in previous studies \cite{botan15,antila19}.
%\todo{Should we comment more the relative quality of different force fields and/or make the subjective force field ranking figures?
%https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/8}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPE.pdf}
  \caption{\label{HGorderParametersPE}
    The headgroup and glycerol backbone order parameters of PE lipids
    from experiments (POPE and signs this work, DPPE from Ref.~\citenum{seelig76}
    and E.coliPE from Ref.~\citenum{gally81}) and simulations with different force fields.
  }
  \todo{This should be clarified as in NMRlipidsI and error bars should be added.
    Probably larger error bars for united atom models based on the report by Fuchs et al.
  } \\
\end{figure}

@@

\section{Lipid headgroups in mixtures with PC and PE or PG lipids}
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of POPC headgroup order parameters with increasing amount of POPE (left) and POPG (right) in bilayer
    from experiments \cite{scherer87,macdonald87} and simulations with different force fields.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{P-N vector angles from Berger-OPLS and GROMOS-CKP simulations are yet to be analyzed.}
\end{figure*}

Headgroup order parameters of PC lipids are unchanged upon addition
of zwitterionic lipids or cholesterol in experiments, but increase
upon addition of negatively charged PG or PS lipids because
headgroup dipole tilts more parallel to the membrane plane after incorporation
of negative charges into the membrane~\cite{seelig87, scherer87,antila18}.
The response of PC headgroup order parameters to the addition of PE or PG lipids
from different simulations is compared with experiments in figure~\ref{HGorderparametersPCvsPEPG}.
by the tested CHARMM36 and
Berger-OPLS force fields, although CHARMM36 slightly overestimates the changes 
The good performance of Berger-OPLS simulations is notable because the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/Höltje force field in our previous work \cite{botan15}.
%\todo{This is text by P. Fuchs, copied from the blog. \\
%Area results in nm$^2$, the error is $<=$ 0.003 nm$^2$ \\
%- pure POPC \\
%CHARMM36: 0.624 \\
%Berger : 0.649 \\
%- POPC/POPE 50:50 \\
%CHARMM36 : POPC 0.609, POPE 0.557 \\
%Berger-hacked: POPC 0.637, POPE 0.632 \\
%----- \\
%One can see that CHARMM 36 predicts a drop in the area on going from pure POPC to POPC/POPE 50:50. This means that POPC pack tightly to POPE. 
%In contrast, the values for Berger are not that changed. 
%The POPE value predicted by CHARMM 36 (in the mixture POPC/POPE 50:50) is much smaller than that predicted by Berger.\\
%-------\\
%The experimental acyl chain order parameters for POPE \cite{pare98} seem larger than reported for POPC \cite{ferreira13},
%which supports the more condensed PE bilayer.
%This is interesting, but not exactly the core message of the manuscript.
%Maybe we should mention this very briefly? For example, we could just report the areas per lipid (without distinguishing PC and PE)
%and mention the difference between CHARMM36 and Berger. I have opened an issue for this: https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/7
%}
